Clusterization Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-10 latex/intro.tex
backup

\subsection{Limitations}\label{subsec:limitations}
The study presented here is not free from drawbacks. First, our
relaxed model does not correspond to any real-life problem. It is
generic enough to be analytically tractable, but one may argue it is
to far removed from any real application. To these claims I would
argue that Mathematics has a long history of studying models that are
bare-bones simplifications of real systems, e.g. the Ising model
\cite{cipra1987} the Lorenz system \cite{brin}, the SIR model
\cite{keeling2008}, to name a few. The second drawback of our model is
that we do not show rigorously that clusterization necessarily occurs,
we only show that it does not pose any obstruction to D-optimality.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{The clusterization effect for the 1D heat equation
    described in Section \ref{subsec:example}. Posterior pointwise
    standard deviations (lines) are plotted over the domain $[0,
      \pi]$, for varying numbers of sensors. Sensor locations
    (circles) were chosen to minimize (an expression analogous to) the
    determinant of the posterior covariance. The clusterization effect
    can be clearly seen for six sensors. Only four observation
    locations are used --- two pairs of sensors are so close they are
    indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

In Section \ref{section:prelim}, we present a more abstract and general
formulation of the inverse problem we consider. But, for the purpose of
illustration, we present clusterization via a toy model --- the 1D heat
equation in $[0,\pi]$ with a homogeneous Dirichlet boundary condition.

@@

The goal is to infer the initial condition $u_0$. For that purpose, we
measure $u$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume centered Gaussian observation
error, so we observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2)$ iid, $\sigma^2 > 0$. We model
the initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. It is
known \cite{Tarantola05} that for linear problems, with Gaussian prior
and error, the posterior is also Gaussian with a covariance that does
not depend on the observed data. The posterior covariance $\postcov$
has a closed form formula, even in infinite dimensions
\cite{Stuart10}. We denote by $\fwd$ the dynamics operator, so that
$u( \cdot,T) = \fwd u_0$, and the observation operator $\obs$ so that
$u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. The posterior covariance is
known and depends only on $\prcov, \fwd, \obs$ and $\sigma^2$ (see
Section \ref{section:prelim} and \eqref{eq:postcov} specifically).

We consider generalization of the information-theoretic design
criterion presented in the introduction to infinite dimensions
(Section \ref{subsec:D optimal design} below). We choose
$\x_j,j=1,\dots,m$ to minimize (an expression analogous to) the
determinant of the posterior covariance operator. We will see later
how this corresponds to maximizing expected information gain.

The clusterization effect is illustrated in Figure
\ref{fig:clusterization illustration}. Posterior pointwise standard
deviations are plotted over the domain $[0, \pi]$. Since the posterior
covariance does not depend on data, the plot has no reference to
actual data observed. The posterior covariance does, however, depend
on location of the observation taken. In Figure
\ref{fig:clusterization illustration}, observation locations are
marked by circles. These were chosen to minimize (an expression
analogous to) the determinant of the posterior covariance. The
clusterization effect can be clearly seen for six sensors. It looks
like only four observations were taken. The reason is that two pairs
of sensors are so close they are indistinguishable.

>>>>

\subsection{Limitations}\label{subsec:limitations}
The study presented here is not free from drawbacks. First, our
relaxed model does not correspond to any real-life problem. It is
generic enough to be analytically tractable, but one may argue it is
to far removed from any real application. To these claims I would
argue that Mathematicians have a long history of studying models that
are bare-bones simplifications of real systems, e.g. the Ising model
\cite{cipra1987} the Lorenz system \cite{brin}, and the SIR model
\cite{keeling2008}, to name a few. The second drawback of our model is
that we do not show rigorously that clusterization necessarily occurs,
we only show that it does not pose any obstruction to D-optimality.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Sensor clusterization for inferring the initial condition
    of the 1D heat equation. Posterior pointwise standard deviations
    (lines) are plotted over the domain $[0, \pi]$, for varying
    numbers of sensors. Sensor locations (circles) were chosen to
    optimize the D-optimality criterion. Sensor clusterization can be
    clearly seen for six sensors. Only four observation locations are
    used --- two pairs of sensors are so close together they are
    indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

For the purpose of illustration, we present clusterization via a toy
inverse problem --- inferring the initial condition of the 1D heat
equation in $[0,\pi]$ with a homogeneous Dirichlet boundary condition.

@@

Our goal here is to infer the initial condition $u_0$. We measure
$u(\x_j, T)$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume iid centered Gaussian observation
error, so we observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2)$ iid, $\sigma^2 > 0$. We model
the initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. It is
known \cite{Tarantola05} that for linear problems, with Gaussian prior
and error, the posterior is also Gaussian with a covariance that does
not depend on the observed data. The posterior covariance $\postcov$
has a closed form formula, even in infinite dimensions
\cite{Stuart10}. We denote by $\fwd$ the forward operator, so that $u(
\cdot,T) = \fwd u_0$, and the observation operator $\obs$ so that
$u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. The posterior covariance is
known and depends only on $\prcov, \fwd, \obs$ and $\sigma^2$ (see
Section \ref{section:prelim} and \eqref{eq:postcov} specifically).

We consider the generalization of the D-optimality criterion to
infinite dimensions \cite{AlexanderianGloorGhattas14} (also, Section
\ref{subsec:D optimal design} below). We choose $\x_j,j=1,\dots,m$ to
minimize the determinant\footnote{Not exactly, see Theorem \ref{thm:}
for a precise statement.} of the posterior covariance operator.

Sensor clusterization is illustrated in Figure \ref{fig:clusterization
  illustration}. Posterior pointwise standard deviations are plotted
over the domain $[0, \pi]$. Since the posterior covariance does not
depend on data, the plot has no reference to actual data observed. The
posterior covariance does, however, depend on location of the
observation taken. Observation locations were chosen to optimize the
D-optimality criterion, and are marked by circles. Sensor
clusterization can be clearly seen for six sensors; it looks like only
four observations were taken. The reason is that two pairs of sensors
are clustered so close together that they are indistinguishable from
one another.

### 2023-04-10 latex/vanishing.tex
backup

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Characterization of D-Optimal Designs with Vanishing Model Error}\label{section:vanishing}
We now turn to analyze D-optimal designs when $\modcov = 0$. Our goal
is to prove Theorem \ref{thm:char} --- a characterization of D-optimal
designs.

%% \subsection{The Nonlinear Eigenvalue Problem}\label{subsec:eigenproblem}
The necessary first-order condition for D-optimality summarized in
Theorem \ref{thm:constrained}, when $\Sigma= \sigma^2I$ (i.e. when
$\modcov = 0$) becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following Proposition and Lemma are proved in the
appendix:

@@

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
We saw that the necessary conditions for a design $\obs$ with $m$
observations to be D-optimal are (1), (2) and (4) of Theorem
\ref{thm:char}. As long as these requirements are satisfied, $\obs$ is
D-optimal. These requirements, however, still leave some freedom in
choosing $\obs$. This freedom may give rise to clusterization, as
illustrated in Figure \ref{fig:clusterization} and explained below.

@@

%% may arise in various scenarios. We first consider
%% sequential designs, and then simultaneous designs.

%% \subsection{Sequential Design}\label{subsec:clusterization sequential}
%% In a sequential optimal design problem, a decision is made on each
%% measurement with all previous measurements fixed. In each step $m=1$,
%% so $k=1$. In this case, Theorem \ref{thm:char} implies each
%% measurement is taken along the direction of the eigenvector of $\fwd
%% \prcov \fwd^*$ with lowest posterior precision. If $\lambda_i$
%% decrease fast enough (e.g. exponentially), then $\lambda_i^{-1}$
%% increase exponentially. Then, at some point, a measurement will be
%% taken along an already measured eigenvector. This situation is
%% measurement clusterization.

%% , $\meas_1 \parallel \ev_1$ and $\eta_1 = 1$. As hinted in
%% section \ref{subsec:seq vs sim}, the posterior becomes the prior for
%% the next step. The eigenvectors of the new posterior are the same as
%% for the previous one. A new observation will be taken in the direction
%% of eigenvector with smalles eigenvalue. This leads to:
%% \begin{align*}
%%   \begin{split}
%%     \meas_2 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_2 \\
%%     \meas_2 \parallel \ev_2  &\text{ o.w. }
%%   \end{split}
%% \end{align*}
%% In the first case, we get clusterization immediately. In the second,
%% we consider $\meas_3$. Same reasoning leads to the conclusion that the
%% eigenvectors of the new prior precision are the same. The eigenvalues,
%% however, are $\lambda^{-1}_1 + \sigma^{-2}, \lambda^{-1}_2 +
%% \sigma^{-2}, \lambda_3^{-1}, \dots$. A new observation is taken in the
%% direction of eigenvector of smallest precision (note that
%% $\lambda^{-1}_1 + \sigma^{-2} < \lambda^{-1}_2 + \sigma^{-2}$):
%% \begin{align*}
%%   \begin{split}
%%     \meas_3 \parallel \ev_1 &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 <
%%     \sigma^2\lambda^{-1}_3 \\
%%     \meas_3 \parallel  \ev_3  &\text{ o.w. }
%%   \end{split}
%% \end{align*}
%% Again, the first case results in clusterization. The sequential design
%% proceeds in this fashion. Every $\meas_i$ is taken in the direction
%% some eigenvector of the prior $\fwd \prcov\fwd^*$. This eigenvector
%% has the smallest eigenvalue of the previous step's posterior precision
%% operator. If we do not encounter clusterization after $m$ such steps,
%% this means that observation $\meas_k$ was taken in the direction
%% $\ev_k$ for $k=1,\dots,m$. Thus:
%% \begin{align}\label{eq:failure}
%%   \begin{split}
%%     \sigma^2\lambda_{k-1}^{-1} + 1 \geq \sigma^2\lambda_k^{-1}, \text{ for } 1\leq k \leq m
%%     \Longrightarrow \lambda_m^{-1} - \lambda_{m-1}^{-1} \leq \sigma^{-2}.
%%   \end{split}
%% \end{align}

%% But \eqref{eq:failure} implies $\{\lambda_i^{-1}\}_{i=1}^{\infty}$
%% grows at most linearly. In section \ref{subsec:abstract OED} we
%% assumed $\fwd$ is strongly smoothing, which means eigenvalues of
%% $(\fwd \prcov\fwd^*)^{-1}$ grow quickly. Thus, if we take $m$ large
%% enough, \eqref{eq:failure} will eventually fail and we will observe
%% sensor clusterization.

%% \subsection{Simultaneous Design}\label{subsec:clusterization simultaneous}
%% If the eigenvalues $\lambda_i$ decrease quickly enough, we will
%% observe clusterizaion.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{D-Optimal Designs with Vanishing Model Error}\label{section:vanishing}
We now turn to analyze D-optimal designs when $\modcov = 0$. Our goal
is to prove Theorem \ref{thm:char} --- a characterization of D-optimal
designs.

The necessary first-order condition for D-optimality summarized in
Theorem \ref{thm:constrained}, when $\Sigma= \sigma^2I$ (i.e. when
$\modcov = 0$) becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following Proposition and Lemma are proved in the
appendix:

@@

\subsection{Clusterization poses no obstruction to D-optimality}\label{subsec:clusterization}
We saw that the necessary conditions for a design $\obs$ with $m$
observations to be D-optimal are (1), (2) and (4) of Theorem
\ref{thm:char}. As long as these requirements are satisfied, $\obs$ is
D-optimal. These requirements, however, still leave some freedom in
choosing $\obs$. This freedom may give rise to clusterization, as
illustrated in Figure \ref{fig:clusterization} and explained below.

@@
